Casual Bandits

### 2016-01-04 TPR_report.tex
Mutiple commits rolled together to fix large file that was accidentally committed.

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}

@@

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

@@

Reinforcement learning addresses the problem of learning from explicitly taking actions. There is typically some state or environment. An agent chooses an action from those available in the current state. The state then evolves stochastically as a function of the selected action and the agent receives some feedback or reward that is a function of the new state. This setting differs from the standard classification problem in that the agent must learn from feedback on the selected action, rather than being presented with the correct action for a given state. A common modelling assumption is that the state evolves only as a function of the previous state and the action chosen, and given these, is independent of the previous history of states and actions. This is known as a Markov decision process or MDP.  A particularly well studied and understood model is MDPs with only a single state. In this case, there are a set of actions, each associated with a fixed but unknown reward distribution and at each time step our agent selects and action and receives corresponding feedback. This is known as the multi-armed bandit problem.

@@

Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process. Causal inference with the problem that full randomisation is not always feasible, affordable or ethical.The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined. The goal of my thesis is to explore these questions. In the next sections I review the key literature in causal inference and bandits. I then present a general approach to  how causal models might be incorporated into bandit settings and conclude by demonstrating an algorithm that leverages causal assumptions to improve performance in a specific bandit setting.

@@

Predicting the outcome of actions without explicitly taking them requires assumptions about how the actions will change the system. A very powerful and general model that underlies much of the recent work in causality is the causal bayesian network. A causal Bayesian network, or directed acyclic graph (DAG), is a Bayesian network in which a link $V_{i} \rightarrow V_{j}$ is defined to mean $V_{i}$ directly causes $V_{j}$. This means that if we intervene and change the value of $V_{i}$, we expect $V_{j}$ to change, but if we intervene to change $V_{j}$, $V_{i}$ will not change. More generally, if $G$ is a causal network for a distribution $P$ defined over variables $V_{1}...V_{N}$, then the distribution after an intervention where we set $X \subset V$ to $x$, denoted $do(X=x)$ is obtained by simply dropping the terms for each of the variables in $X$ from the factorization given by the network. This is referred to as the truncated product formula \cite{Pearl2000}.

If the network contains latent (unobservable) variables we will not be able to calculate all the terms in the truncated product formula. However, it may still be possible to determine the post-interventional distribution of specific variables of interest. A general causal query $P(Y|do(X=x))$ is identifiable if it can be shown to be equivalent to an expression containing only pre-interventional quantities. This means that, asymptotically, we obtain an unbiased estimate for the distribution after an intervention based on purely observational data.

The do calculus is a set of three rules that allow transformations of interventional terms to non-interventional ones, given a causal graph \cite{Pearl2000}. They are derived directly from d-seperation properties of graphical models and the definition of intervention in causal DAGs. These rules are complete. A query is identifiable if and only if it can be transformed to contain only non-interventional terms via the do-calculus \cite{Shpitser2006,Huang2006} and there is an equivalent algorithm that can take any causal graph and query and determine identifiability \cite {Shpitser2008} (see XXXX for a javascript demonstration of this algorithm). If a query is not identifiable, it may still be possible to get bounds for causal effects, for example using instrumental variables \cite{Angrist2009} or by making additional assumptions.

@@

There are differences between the models we have considered when it comes to these non-interventional queries. Counterfactuals are not defined in causal Bayesian networks, as they only encode information on the interventional distribution over variables.  Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made. Structural equation models with independent errors allow the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.

In practice, differences in focus and approach across different fields eclipse the actual differences in the models. The work on causal graphical models \cite{Pearl2000,Sprites} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations. The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment. This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}. In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field. Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}. In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.

@@

If the end goal of causal discovery is to estimate causal effects, then it may not be necessary to learn the entire graph, only the subset of the graph surrounding target variables of interest. Such local causal discovery techniques can be scaled to problems with many more variables \cite{Aliferis2002}. Once a set of causal graphs has been identified, causal effects of interest can be bounded by combining the results for the all the networks. This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}. An implementation is available in the R package \cite{Kalisch2012}

@@

Finally, rather than explicitly developing an algorithm based on a specific asymmetry between cause and effect, \cite{LopezPaz2014} propose learning what causality looks like from data. They take as input a dataset where each row of data is itself a dataset in which either $X \rightarrow Y$ or $Y \rightarrow X$ and a corresponding label. Estimates of the distributions $P(X)$, $P(Y)$ and $P()$ for each row are then mapped to features in some kernel space via mean kernel embeddings and finally a standard classification algorithm can be trained to learn label. New datasets where the direction of causality is unknown are then simply mapped to the kernel space and the causal direction is classified according to the trained classifier. In practice, the classifier is trained mostly on simulated data, as it is difficult to find a sufficient set of causal problems with only two variables where the direction of causality is known.

@@

In its classic formulation \cite{Robbins1952} the (stochastic) K-armed bandit describes a sequential decision making problem, with $K$ possible actions or arms. Each arm $i$ is associated with a fixed but unknown reward distribution $\nu_i$. For each timestep $t$ upto a horizon $T$ the learner selects an action $I_t \in \set{1...K}$ and receives a reward, $g_{I_t,t}$, sampled i.i.d from $\nu_i$. The goal of the learner is to maximize the total reward they receive. This problem introduces the fundamental exploration-exploitation trade-off. The learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.

@@

The classic multi-armed bandit is a powerful tool for sequential decision making. However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces. This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example linear bandits, dependent bandits, X-armed bandits and gaussian process bandits, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.

@@

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}. This is an expansion of the model shown in figure \ref{fig:unify_frameworks}. We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.

@@

We then simply play the arm with the highest estimated reward for the remaining $T-h$ timesteps. We can show that we can select an $h$ (based on $T$, $N$ and $m$) that leads to worst case regret $R_T = \bigotilde{T^{2/3}m^{1/3}}$. This result can still be obtained in the case where $\boldsymbol{q}$ is unknown and must be learned during the exploration time. Comparing this to the standard bandit regret, $R_T = \bigo{\sqrt{TK}}$, we expect the causal algorithm to do better if $m < \frac{K^{3/2}}{\sqrt{T}}$. \footnote{ $\bigotilde{.}$ and $\bigthetatilde{.}$ suppress polylogarithmic factors}

In this algorithm, we have only used the side information provided by the $do()$ action about other actions. Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen. The feedback graph is distinct from the causal graph. A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$. For this specific problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions. The Exp3-IX algorithm \cite{Kocak2014} was developed for the adversarial version of this problem and has regret $\bigo{\sqrt{\bar{\alpha}T}}$, where $\bar{\alpha}$ is the average independence number of $G_t$. In our case $\bar{\alpha} = \frac{N}{2}$ so we again obtain the regret of the standard bandit algorithm. The issue here is that a malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action. This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$

@@

>>>>

\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}

@@

\DeclareMathOperator{\ringarrow}{\raisebox{0.5ex}{\tikz[baseline]{\draw[ring->](0,0)--(2em,0);}}}
\DeclareUnicodeCharacter{00A0}{ }
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

@@

Reinforcement learning addresses the problem of learning from explicitly taking actions. There is typically some state or environment. An agent chooses an action from those available in the current state. The state then evolves stochastically as a function of the selected action and the agent receives some feedback or reward that is a function of the new state. This setting differs from the standard classification problem in that the agent must learn from feedback on the selected action, rather than being presented with the correct action for a given state. A common modelling assumption is that the state evolves only as a function of the previous state and the action chosen, and given these, is independent of the previous history of states and actions. This is known as a Markov decision process or MDP.  A particularly well studied and understood model is the singe state MDP. In this case, there are a set of actions, each associated with a fixed but unknown reward distribution and at each time step our agent selects an action and receives corresponding feedback. This is known as the multi-armed bandit problem.

@@

Both approaches can be seen as extensions to the concept of randomised controlled trials. Bandit algorithms deal with the sequential nature of the decision making process, causal inference with the problem that full randomisation is not always feasible, affordable or ethical. The similarities between the problems that these techniques have been developed to address raises the question of if there are problems best addressed by a combination of these approaches and how they can be combined. The goal of my thesis is to explore these questions. In the next sections I review the key literature in causal inference and bandits. I then present a general approach to  how causal models might be incorporated into bandit settings and conclude by demonstrating an algorithm that leverages causal assumptions to improve performance in a specific bandit setting.

@@

Predicting the outcome of actions without explicitly taking them requires assumptions about how the actions will change the system. A very powerful and general model that underlies much of the recent work in causality is the causal bayesian network. A causal bayesian network, or directed acyclic graph (DAG), is a bayesian network in which a link $V_{i} \rightarrow V_{j}$ is defined to mean $V_{i}$ directly causes $V_{j}$. This means that if we intervene and change the value of $V_{i}$, we expect $V_{j}$ to change, but if we intervene to change $V_{j}$, $V_{i}$ will not change. More generally, if $G$ is a causal network for a distribution $P$ defined over variables $V_{1}...V_{N}$, then the distribution after an intervention where we set $X \subset V$ to $x$, denoted $do(X=x)$ is obtained by simply dropping the terms for each of the variables in $X$ from the factorization given by the network. This is referred to as the truncated product formula \cite{Pearl2000}.

If the network contains latent (unobservable) variables we will not be able to calculate all the terms in the truncated product formula. However, it may still be possible to determine the post-interventional distribution of specific variables of interest. A general causal query $P(Y|do(X=x))$ is identifiable if it can be shown to be equivalent to an expression containing only pre-interventional quantities. This means that, asymptotically, we can obtain an unbiased estimate for the distribution after an intervention based on purely observational data.

The do calculus is a set of three rules that allow transformations of interventional terms to non-interventional ones, given a causal graph \cite{Pearl2000}. They are derived directly from d-seperation properties of graphical models and the definition of intervention in causal DAGs. These rules are complete. A query is identifiable if and only if it can be transformed to contain only non-interventional terms via the do-calculus \cite{Shpitser2006,Huang2006}. There is an equivalent algorithm that can take any causal graph and query and determine identifiability \cite {Shpitser2008} (see \url{http://finnhacks42.github.io/causal_identify} for a javascript demonstration of this algorithm). If a query is not identifiable, it may still be possible to get bounds for causal effects, for example using instrumental variables \cite{Angrist2009} or by making additional assumptions.

@@

There are differences between the models we have considered when it comes to these non-interventional queries. Counterfactuals are not defined in causal bayesian networks, as they only encode information on the interventional distribution over variables.  Counterfactuals can be defined in terms of structural equation models \cite{Pearl2000} but there are subtle differences depending on the form of assumptions made. Structural equation models with independent errors allow the identification of quantities in mediation studies, which are not identifiable with the weak ignorability assumptions and cannot be tested experimentally \cite{Richardson2013}.

In practice, differences in focus and approach across different fields eclipse the actual differences in the models. The work on causal graphical models \cite{Pearl2000,Sprites2000} focuses on non-parametric estimation in the population limit and rigorous theoretical foundations. The Neyman-Rubin framework builds on our understanding of randomized experiment and generalizes to quasi-exeperimental and observational settings, with a particular focus on non-random assignment to treatment. This research emphasises estimating average causal effects and provides practical methods for estimation, in particular, propensity scores; a method to control for multiple variables in high dimensional settings with finite data \cite{Rosenbaum1983}. In economics, inferring causal effects from non-experimental data so as to support policy decisions is central to the field. Economists are often interested in broader measures of the distribution of causal effects than the mean and make extensive use of structural equation models, generally with strong parametric assumptions \cite{Heckman2008}. In addition, the parametric structural equation models favoured in economics can be extended to analyse cyclic (otherwise referred to as non-recursive) models.

@@

If the end goal of causal discovery is to estimate causal effects, then it may not be necessary to learn the entire graph, only the subset of the graph surrounding target variables of interest. Such local causal discovery techniques can be scaled to problems with many more variables \cite{Aliferis2002}. Once a set of causal graphs has been identified, causal effects of interest can be bounded by combining the results for the all the networks. This procedure is the IDA algorithm \cite{Maathuis2009} and has been found to outperform standard regularization techniques at finding causal effects in a high-dimensional yeast gene expression data set \cite{Maathuis2010}. An implementation is available in the R package pcalg \cite{Kalisch2012}

@@

Finally, rather than explicitly developing an algorithm based on a specific asymmetry between cause and effect, \cite{LopezPaz2014} propose learning what causality looks like from data. They take as input a dataset where each row of data is itself a dataset in which either $X \rightarrow Y$ or $Y \rightarrow X$ and a corresponding label. Estimates of the distributions $P(X)$, $P(Y)$ and $P(X,Y)$ for each row are then mapped to features in some kernel space via mean kernel embeddings and finally a standard classification algorithm can be trained to learn the labels. New datasets, where the direction of causality is unknown, are then simply mapped to the kernel space and the causal direction is classified according to the trained classifier. In practice, the classifier is trained mostly on simulated data as it is difficult to find a sufficient set of causal problems with only two variables, where the direction of causality is known.

@@

In its classic formulation \cite{Robbins1952} the (stochastic) K-armed bandit describes a sequential decision making problem, with $K$ possible actions or arms. Each arm $i$ is associated with a fixed but unknown reward distribution $\nu_i$ \footnote{In order to obtain regret bounds, some assumptions are required on the distributions $\nu_i$. It sufficient to assume they are sub-gaussian}.  For each timestep $t$ upto a horizon $T$ the learner selects an action $I_t \in \set{1...K}$ and receives a reward, $g_{I_t,t}$, sampled i.i.d from $\nu_i$. The goal of the learner is to maximize the total reward they receive. This problem introduces the fundamental exploration-exploitation trade-off. The learner must balance playing arms that have yielded good results previously with exploring arms about which they are uncertain.

@@

Another problem that has attracted a lot of recent attention \cite{Bubeck2009,Audibert2010,Gabillon2012,Karnin2013} within the multi-armed bandit framework is \textit{pure exploration} or \textit{best arm identification}. In this setting, the horizon $T$ represents a fixed budget for exploration after which the algorithm outputs a single best arm $i$. The performance of the algorithm is measured by the simple regret; the expected difference between the mean reward of the (truly) optimal arm and the mean reward of the arm selected by the algorithm, $R_s(T) = \mu^* - \E{\mu_i}$ . This problem arises naturally in applications where there is a testing or evaluation phase, during which regret is not incurred, followed by a commercialization or exploitation phase. For example, many strategies might be assessed via simulation prior to one being selected and deployed. The simple regret for a K-armed bandit is lower bounded by $\bigo{\sqrt{K/T}}$ \cite{Bubeck2009}.

The classic multi-armed bandit is a powerful tool for sequential decision making. However, the regret grows linearly with the number of (sub-optimal) actions and many real world problems have large or even infinite action spaces. This has led to the development of a wide range of models that assume some structure across the reward distributions for different arms, for example generalized linear bandits \cite{filippi2010parametric}, dependent bandits \cite{Pandey2007}, X-armed bandits \cite{Bubeck2010} and gaussian process bandits \cite{srinivas2009gaussian}, or that consider more complex feedback, for example the recent work on graph feedback \cite{Mannor2011,Lelarge2012,Alon2013,Buccapatnam2014,Kocak2014,Alon2015} and partial monitoring \cite{Piccolboni2001,Bartok2014}.

@@

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}. This is an expansion of the model shown in figure \ref{fig:unify_frameworks}. We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.

@@

We then simply play the arm with the highest estimated reward for the remaining $T-h$ timesteps. We can show that we can select an $h$ (based on $T$, $N$ and $m$) that leads to worst case regret $R_T = \bigotilde{T^{2/3}m^{1/3}}$. This result can still be obtained in the case where $\boldsymbol{q}$ is unknown and must be learned during the exploration time. Comparing this to the standard bandit regret, $R_T = \bigo{\sqrt{TK}}$, we expect the causal algorithm to do better if $m < \frac{K^{3/2}}{\sqrt{T}}$. We can also show that the simple regret $R_s \in \bigotilde{\sqrt{m/T}}$. If $m << K$, this represents a significant improvement over standard bandit setting. \footnote{ $\bigotilde{.}$ and $\bigthetatilde{.}$ suppress polylogarithmic factors}

In this algorithm, we have only used the side information provided by the $do()$ action about other actions. Since the $do()$ action fully reveals the value of alternate actions we could have incorporated this information via the graph feedback model \cite{Mannor2011}, where at each timestep the feedback graph $G_t$ is selected stochastically, dependent on $\boldsymbol{q}$, and revealed after an action has been chosen. The feedback graph is distinct from the causal graph. A link $A \rightarrow B$ in $G_t$ indicates that selecting the action $A$ reveals the reward for action $B$. For this specific problem, $G_t$ will always be a star graph with the action $do()$ connected to half the remaining actions. The Exp3-IX algorithm \cite{Kocak2014} was developed for the adversarial version of this problem and has regret $\bigo{\sqrt{\bar{\alpha}T}}$, where $\bar{\alpha}$ is the average independence number of $G_t$. In our case $\bar{\alpha} = \frac{N}{2}$ so we again obtain the regret of the standard bandit algorithm. The issue here is that a malicious adversary can select the same graph each time, such that the rewards for half the arms are never revealed by the informative action. This is equivalent to a, nominally, stochastic selection of feedback graph where $\boldsymbol{q} = \boldsymbol{0}$

@@

\section{Paid work statement}
I am not currently undertaking or planning to undertake any paid work.

### 2016-01-04 causal-bandit/icml-2015/causal-bandits-icml.tex
Mutiple commits rolled together to fix large file that was accidentally committed.

\usepackage{tikz}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{cleveref}
\usetikzlibrary{arrows,positioning} 
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.5em,
           minimum height=2.5em,
           font=\footnotesize,
           text centered,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

@@

@@

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}. We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$, or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.

@@

\eqn {
\label{eq:estimation_transfer}
P(Y_t|do(X_{i,t} = j))=\\
 \sum_{j'} P(Y_t|do(X_{l,t} = j'),X_{i,t} = j)P(X_{l_t} = j')
}

@@

Algorithm \ref{alg:known_q} tries to learn the rewards for all the arms during an exploration phase $h$ and then picks the arm with the highest empirical mean for all remaining timesteps. During its exploration phase, it learns all the frequently occurring actions by observation and the remaining, infrequently occurring actions, by explicitly playing them. This leads to Chernoff type high probability bounds on the difference between the empirical and true rewards for all arms of the form $P(\hat{\mu}_{i,j} - \mu_{i,j} > D) \leq e^{-hD^2/m}$. By choosing optimal values for $D$ and $h$ we obtain the regret bound given in \cref{thm:known_q_regret}. A full proof is given in the supplementary materials. (\color{red} or should we put it here? \color{black})

@@

\section{Related Work}
There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures. Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc. There is also an active line of research into bandits with feedback defined by a graph. Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action []. The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.

@@

\section{Conclusion}

>>>>

\usepackage{tikz}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{cleveref}

\graphicspath{ {figures/} }
\usetikzlibrary{arrows,positioning} 
\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.5em,
           minimum height=2.5em,
           font=\footnotesize,
           text centered,
           fill=blue!20!white},
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }  
}

@@

There has been substantial recent work into extending bandit algorithms to incorporate additional assumptions and deal with more complex feedback structures. Algorithms with strong guarantees have been developed for linear bandits [], generalized linear bandits, gaussian process bandits [], etc. There is also an active line of research into bandits with feedback defined by a graph. Actions are modelled as nodes in the graph and the agent observes rewards for each action connected to the selected action []. The novelty of our work is that we assume prior knowledge of the causal structure but not the functional form of the relationship between variables.

@@

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{N}\}$ that independently cause a target variable of interest $Y$, figure \ref{fig:causalStructure}. We can run sequential experiments on the system, where at each timestep $t$ we can either do nothing, $do()$ \footnote{A note on notation: In the bandit community it is implicit that algorithms selecting actions are intervening in the system. So it is sufficient to index actions according to the variable and value. However, in causal graphs, it is essential to differentiate observing (or conditioning) on a variable taking a certain value, from intervening to set that variable. Although in the specific causal graph we consider, observation and intervention are the same, we deliberately introduce the do-notation \cite{Pearl2000} that makes this distinction clear so as to help provide a bridge between the bandit and causal inference communities.} , or select a single variable, $X_i$, on which to intervene, $do(X_{t,i} = J_t)$, and subsequently observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$.

@@

\eqn {
\label{eq:estimation_transfer}
P(Y_t|do(X_{i,t} = j))= & \nonumber \\
 \sum_{j'}  P(Y_t|do(X_{l,t} & = j'),X_{i,t} = j)P(X_{l_t} = j')
}

@@

Algorithm \ref{alg:known_q} tries to learn the rewards for all the arms during an exploration phase $h$ and then picks the arm with the highest empirical mean for all remaining timesteps. During its exploration phase, it learns all the frequently occurring actions by observation and the remaining, infrequently occurring actions, by explicitly playing them. This leads to Chernoff type high probability bounds on the difference between the empirical and true rewards for all arms of the form $P(\hat{\mu}_{i,j} - \mu_{i,j} > D) \leq e^{-hD^2/m}$. By choosing optimal values for $D$ and $h$ we obtain the regret bound given in \cref{thm:known_q_regret}. A full proof is given in the supplementary materials.

@@

\begin{figure}
\caption{Final regret versus number of variables $N$ for UCB with $\alpha = 2$, Causal-Explore-Exploit with $m=2$ and with $m=N$ and horizon $T = 10,000$ . Error bars show standard deviation over 100 simulations. The regret for UCB grows linearly with the number of variables, whist for Causal-Explore-Exploit with fixed $m$, the growth is sub-logarthmic.  }
\label{fig:known_q_r_vs_N}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_N_T10000_sims100_20151229_113550.pdf}
\end{figure}

\begin{figure}
\caption{Cumulative regret over time for $N = 17$ for UCB with $\alpha=2$, Causal-Explore-Exploit with $m=2$ and Causal-Explore-Exploit with $m=N$. Shaded region shows standard deviation over 100 simulations. The Causal-Explore-Exploit algorithm incurs linear regret during the exploration phase, after which it selects the optimal arm with high probability. For $m=2$, we have $K \sim m^{2/3}T^{1/3}$ and see that we are in the regime in which Causal-Explore-Exploit outperforms UCB.}
\label{fig:known_q_r_vs_t}
\centering
\includegraphics[width=.5\textwidth]{exp_regret_vs_t_T10000_N17_sims100_20151229_120647.pdf}
\end{figure}

\section{Discussion}

@@

\section{Future Open Questions}
\begin{itemize}
\item Known but arbitrary structure
\item Learning structure then exploiting
\end{itemize}
\section{Conclusion}
